---
title: 'FDOS Group Project'
subtitle: 'Subtitle'
author: "Miraç Arda Balaban, Rachel Kane, Lucas Mordue, Gretchen Moulton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    df-print: paged
    fig-width: 9
    fig-height: 6
    page-layout: full
---

```{r}
# Install and load required packages
if (!require("pacman")) install.packages("pacman")
library(pacman)

p_load(readxl, dplyr, ggplot2, knitr, lubridate, tidyr, sf, httr)
```

# Introduction

Include:

* Research Question & target variable for prediction

* Motivation/relevance - cite some literature

* Dataset name 

* Setting description 

* Relevant features for prediction in the dataset

* Main methods used

* Key findings

# Data Description

The data source is the [New York Police Department Stop, Question and Frisk Data](https://www.nyc.gov/site/nypd/stats/reports-analysis/stopfrisk.page).

* Dimension

* How data is collected & what universe is/is not observed

* Temporal and spatial span

First, we set up file paths and import the data.

```{r}
# Define base file path - change as needed
wd <- "C:/Users/rache/OneDrive/Documents/GitHub/bse-masters/foundations-of-data-science/Foundations of Data Science"

# Define input file path
input_folder <- file.path(wd, "input")

# Set working directory
setwd(wd)

# Import the data
sqf_data <- read_excel(file.path(input_folder, "sqf-2023.xlsx"))

 #View the first few rows of the data
head(sqf_data)

# Check original dimensions
dim(sqf_data)

# View head
head(sqf_data)
```

# Data Cleaning

## Column Names

First, we change column names from strictly upper case to strictly lower case, because it's cuter.

```{r}
colnames(sqf_data) <- tolower(colnames(sqf_data))

# check
colnames(sqf_data)[1:3]
```

## Missing Values 

There are many observations in the data with values == `(null)` across different columns.

The overall number of `(null)` values are as follows:

```{r}
# Sum (null) values by column
apply(sqf_data, 2, function(x) sum(x == "(null)"))
```

Note, however, that not all of these `(null)` observations are equivalent:

* in some columns, `(null)` means the data are **genuinely effectively `NA`**, as there are instances of both "Y" and "N" (for binary variable for example), alongside `(null)`.

```{r}
sqf_data %>% 
  group_by(ask_for_consent_flg) %>% 
  summarise(N = n()) %>% 
  kable()
```

* whereas in other cases, the `null` values are, confusingly, actually used as "N".

```{r}
print(unique(sqf_data$firearm_flag))

# grouped example
sqf_data %>% 
  group_by(weapon_found_flag, firearm_flag) %>% 
  summarise(N = n()) %>% 
  kable()
```

note issue here re no weapons found but firearm found for small n of cases

We deal with these cases separately below:

```{r}
# columns with genuinely missing null values
type1_variables <- c("ask_for_consent_flg", "consent_given_flg", "suspect_reported_age", "suspect_height", "suspect_weight", "suspect_arrest_offense") 

# pre-clean check
print(unique(sqf_data$ask_for_consent_flg))

# replace these with NAs
sqf_data <- sqf_data %>%
  mutate(across(all_of(type1_variables), ~ ifelse(. == "(null)", NA, .)))

# post-clean check
print(unique(sqf_data$ask_for_consent_flg))

# columns with null values meaning N
type2_variables <- c("firearm_flag", "knife_cutter_flag", "other_weapon_flag") 

# pre-clean check
print(unique(sqf_data$firearm_flag))

# replace these with NAs
sqf_data <- sqf_data %>%
  mutate(across(all_of(type2_variables), ~ ifelse(. == "(null)", "N", .)))

# post-clean check
print(unique(sqf_data$firearm_flag))
```
  
### Errors in Weapon Found Flag

For now, replace `"N"` values of `weapon_found_flag` with y if any of the weapons categories are `"Y"`

```{r}
print(sum(sqf_data$weapon_found_flag == "N"))

sqf_data <- sqf_data %>% 
  mutate(weapon_found_flag = ifelse((firearm_flag == "Y" | knife_cutter_flag == "Y" | other_weapon_flag == "Y"), "Y", weapon_found_flag))

print(sum(sqf_data$weapon_found_flag == "N"))
```

  
## Binary Variables 
  
can replace Ys and Ns here with 1s and 0s


# Exploratory Data Analysis 

## Frisk Status

```{r}
# frisked
sqf_data %>% 
  group_by(frisked_flag) %>% 
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>% 
  arrange(desc(N)) %>% 
  kable(booktabs = TRUE, col.names = c("Suspect Frisked", "N Stops", "% Total Stops"), align = "l")

# by suspected crime
sqf_data %>% 
  group_by(suspected_crime_description) %>%
  summarize(N = n(),
            pc_frisked = round((sum(frisked_flag == "Y")/N) * 100, 2),
            pc_not_frisked = round((sum(frisked_flag == "N")/N) * 100, 2)
            ) %>% 
  arrange(desc(pc_frisked)) %>% 
  kable(booktabs = TRUE, col.names = c("Suspected Crime", "N", "% Frisked", "% Not Frisked"), align = "l")

# by race
sqf_data %>% 
  group_by(suspect_race_description) %>%
  summarize(N = n(),
            pc_frisked = round((sum(frisked_flag == "Y")/N) * 100, 2),
            pc_not_frisked = round((sum(frisked_flag == "N")/N) * 100, 2)
            ) %>% 
  arrange(desc(pc_frisked)) %>% 
  kable(booktabs = TRUE, col.names = c("Suspect Race", "N", "% Frisked", "% Not Frisked"), align = "l")

```

## Search Status

```{r}
# searched
sqf_data %>%
  group_by(searched_flag) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Suspect Searched", "N Stops", "% Total Stops"), align = "l")
```

## Arrest Status

```{r}
sqf_data %>%
  group_by(suspect_arrested_flag) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Suspect Arrested", "N Stops", "% Total Stops"), align = "l")

# plot
ggplot(data = sqf_data, aes(x = suspect_arrested_flag, fill = suspect_race_description)) +
  geom_bar() +
  coord_flip() +
  theme_minimal() +
  xlab("Suspect Arrested") +
  ylab("Number of Observations") +
  scale_fill_brewer(type = "qual", palette = "Spectral", name = "Suspect Race") +
  labs(title = "Suspect Arrest Status by Race")

# 100% stacked version
ggplot(data = sqf_data, aes(x = suspect_arrested_flag, fill = suspect_race_description)) +
  geom_bar(position = "fill") +
  coord_flip() +
  theme_minimal() +
  xlab("Suspect Arrested") +
  ylab("% Observations") +
  scale_fill_brewer(type = "qual", palette = "Spectral", name = "Suspect Race") +
  labs(title = "Suspect Arrest Status by Race")

```

## Physical Force Status

```{r}
```

## Other Univariate Descriptives

```{r}
sqf_data %>%
  group_by(suspected_crime_description) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Suspected Crime", "N Stops", "% Total Stops"), align = "l")


# # suspect sex
sqf_data %>%
  group_by(suspect_sex) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Suspect Sex", "N Stops", "% Total Stops"), align = "l")  

# suspect race
sqf_data %>%
  group_by(suspect_race_description) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Suspect Race", "N Stops", "% Total Stops"), align = "l")  # small amount missing - solution?

# issuing office rank
sqf_data %>%
  group_by(issuing_officer_rank) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Issuing Officer Rank", "N Stops", "% Total Stops"), align = "l")  


# supervising office rank
sqf_data %>%
  group_by(supervising_officer_rank) %>%
  summarise(N = n(),
            Pc = N / nrow(sqf_data) * 100) %>%
  arrange(desc(N)) %>%
  kable(booktabs = TRUE, col.names = c("Supervising Officer Rank", "N Stops", "% Total Stops"), align = "l")
```

## Weapon Possession

```{r}
sqf_data <- sqf_data %>%
  mutate(weapon_type = case_when(
    firearm_flag == "Y" & knife_cutter_flag == "Y" & other_weapon_flag == "Y" ~ "Firearm, Knife Cutter, and Other Weapon",
    firearm_flag == "Y" & knife_cutter_flag == "Y" ~ "Firearm and Knife Cutter",
    firearm_flag == "Y" & other_weapon_flag == "Y" ~ "Firearm and Other Weapon",
    knife_cutter_flag == "Y" & other_weapon_flag == "Y" ~ "Knife Cutter and Other Weapon",
    firearm_flag == "Y" ~ "Firearm Only",
    knife_cutter_flag == "Y" ~ "Knife Cutter Only",
    other_weapon_flag == "Y" ~ "Other Weapon Only",
    TRUE ~ "No Weapon Found"
  ))

ggplot(sqf_data, aes(x = weapon_found_flag, fill = weapon_type)) +
  geom_bar() +
  labs(
    title = "Weapons Found by Type of Weapon(s)",
    x = "Type of Weapon(s) Found",
    y = "Number of Cases"
  ) +
  theme_minimal() +
  scale_fill_brewer(type = "qual", palette = "Spectral", name = "Weapon Type")
```


* basic grouped tables and plots, cumulative plots

* correlation plots





# Regression Analysis

* state models, assumptions, evaluation criteria blah blah blah
* run + check model diagnostics

# Conclusion
